---
title: "Summary report"
author: authername
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
```

```{r load_library}
renv::load(Sys.getenv("PROJECT_DIR"))
library(reshape2)
library(ggplot2)
library(dplyr)
library(grid)
library(gridExtra)
library(openxlsx)
library(BPCells)
library(Seurat)
library(scplotter)
library(RColorBrewer)
library(patchwork)
default_width <- 6
default_height <- 6
```

# Sample information

```{r data_load}
opt <- params$opt
sampleinfo <- read.delim(opt$samplesheet, header = T,check.names=F,stringsAsFactors=F)
seurat_obj <- readRDS(opt$seurat_obj)
DT::datatable(sampleinfo,rownames = F)
```

# Cluster summary {.tabset}

## Number of cells per cluster

```{r number_of_cells_table}
cell_per_cluster <- as.data.frame(unclass(table(seurat_obj$seurat_clusters,seurat_obj$condition)))
cell_per_cluster <- cbind(data.frame(cluster_number = rownames(cell_per_cluster)), cell_per_cluster)
write.table(cell_per_cluster,file="./cellNumberbyClusterbyCondition.txt",row.names = F,col.names=T, sep="\t",quote=FALSE)
cell_per_cluster_pct <- cell_per_cluster
cell_per_cluster_pct[-1] <- sweep(cell_per_cluster_pct[-1],2,apply(cell_per_cluster_pct[-1], 2,sum),"/")
write.table(cell_per_cluster_pct,file="./cellPctbyClusterbyCondition.txt",row.names = F,col.names=T, sep="\t",quote=FALSE)

cell_per_cluster_bysample <- as.data.frame(unclass(table(seurat_obj$seurat_clusters,seurat_obj$sampleid)))
cell_per_cluster_bysample <- cbind(data.frame(cluster_number = rownames(cell_per_cluster_bysample)), cell_per_cluster_bysample)
write.table(cell_per_cluster_bysample,file="./cellNumberbyClusterbySampleid.txt",row.names = F,col.names=T, sep="\t",quote=FALSE)
cell_per_cluster_bysample_pct <- cell_per_cluster_bysample
cell_per_cluster_bysample_pct[-1] <- sweep(cell_per_cluster_bysample_pct[-1],2,apply(cell_per_cluster_bysample_pct[-1], 2,sum),"/")
write.table(cell_per_cluster_bysample_pct,file="./cellPctbyClusterbySampleid.txt",row.names = F,col.names=T, sep="\t",quote=FALSE)

DT::datatable(cell_per_cluster,rownames = F)
```

## Number of cells per cluster bar plot

```{r number_of_cells_barplot}
temp <- melt(table(seurat_obj$seurat_clusters,seurat_obj$condition))
temp <- temp %>% group_by(Var2) %>% mutate(pct=value/sum(value))

gp <- CellStatPlot(seurat_obj, group_by = "condition", palette = "Set2") + labs(x='cluster number', y='number of cells') + theme(axis.text = element_text(size=15), axis.title = element_text(size=18,face="bold"))

ggsave('./numberofcellbarplot.png',gp,width=12,height=12, dpi = 300)
ggsave('./numberofcellbarplot.pdf',gp,width=12,height=12, useDingbats=FALSE)
gp
```

## Percentage of cells per cluster bar plot

```{r number_of_cells_barplot_pct}
gp <- CellStatPlot(seurat_obj, group_by = "condition", palette = "Set2", frac = "group") + labs(x='cluster number', y='percentage of cells') + theme(axis.text = element_text(size=15), axis.title = element_text(size=18,face="bold"))

ggsave('./pctofcellbarplot.png',gp,width=12,height=12,dpi = 300)
ggsave('./pctofcellbarplot.pdf',gp,width=12,height=12, useDingbats=F)
gp
```

## Percentage of cells per cluster pie chart

```{r, echo=FALSE}
piechart_width <- 8*ceiling(length(unique(temp$Var2))/round(sqrt(length(unique(temp$Var2)))))
piechart_height <- 6*round(sqrt(length(unique(temp$Var2))))
```

```{r number_of_cells_piechart_pct, fig.width=piechart_width, fig.height=piechart_height}
gg_color <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

cellpct_piechart <- function(){
  par(mfrow=c(round(sqrt(length(unique(temp$Var2)))),ceiling(length(unique(temp$Var2))/round(sqrt(length(unique(temp$Var2)))))))
for(i in unique(temp$Var2)){
  pielabels <- sprintf("%s: %3.1f%s", paste0('cluster',temp$Var1[temp$Var2 == i]),100*temp$pct[temp$Var2 ==i], "%")  
  pie(temp$pct[temp$Var2 ==i],
      labels=pielabels,
      clockwise=TRUE,
      radius=1,
      col=gg_color(18),
      border="white",
      cex=1,
      main=i)
}
}

png('./pctofcellpiechart.png',res = 300,width = piechart_width,units = 'in',height = piechart_height)
cellpct_piechart()
invisible(dev.off())

pdf('./pctofcellpiechart.pdf',width =piechart_width ,height =piechart_height)
cellpct_piechart()
invisible(dev.off())

## generate plot
cellpct_piechart()
par(mfrow = c(1,1))
```

# {-}

# Dimensionality reduction visualization {.tabset}

## Overall

```{r dim_reduc_overall}
if(opt$sketch_flag == "1") seurat_obj <- subset(seurat_obj, cells = colnames(seurat_obj@assays$sketch))
p <- CellDimPlot(seurat_obj,group_by = "seurat_clusters", reduction = opt$vismethod, theme = "theme_blank", highlight =TRUE, label = TRUE, label_insitu = TRUE)
ggsave(paste0('./',opt$vismethod,'.png'),p,width=6,height=5, dpi = 300)
ggsave(paste0('./',opt$vismethod,'.pdf'),p,width=6,height=5, useDingbats=FALSE)
p
```

## Split by sample

```{r, echo=FALSE}
nsamples <- length(unique(seurat_obj$sampleid))
dimplot_width <- 6*round(sqrt(nsamples))
dimplot_height <- 6*ceiling(nsamples/round(sqrt(nsamples)))
```

```{r dim_reduc_sample, fig.width=dimplot_width, fig.height=dimplot_height}
p <- CellDimPlot(seurat_obj,group_by = "seurat_clusters", reduction = opt$vismethod, theme = "theme_blank", highlight =TRUE, label = TRUE, label_insitu = TRUE, split_by = "sampleid", ncol = round(sqrt(nsamples)))
ggsave(paste0('./',opt$vismethod,'_split_sample.png'),p,width=dimplot_width,height=dimplot_height,limitsize = FALSE,dpi=300) 
ggsave(paste0('./',opt$vismethod,'_split_sample.pdf'),p,width=dimplot_width,height=dimplot_height,limitsize = FALSE,useDingbats=FALSE)
p
```

## Split by condition

```{r, echo=FALSE}
nconditions <- length(unique(seurat_obj$condition))
dimplot_width_condition <- 6*round(sqrt(nconditions))
dimplot_height_condition <- 6*ceiling(nconditions/round(sqrt(nconditions)))
```

```{r dim_reduc_condition, fig.width=dimplot_width_condition, fig.height=dimplot_height_condition}
p <- CellDimPlot(seurat_obj,group_by = "seurat_clusters", reduction = opt$vismethod, theme = "theme_blank", highlight =TRUE, label = TRUE, label_insitu = TRUE, split_by = "condition", ncol = round(sqrt(nconditions)))
ggsave(paste0("./",opt$vismethod,'_split_condition.png'),p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,dpi=300) 
ggsave(paste0("./",opt$vismethod,'_split_condition.pdf'),p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,useDingbats=FALSE)
p
```

## Spatial plot

```{r, echo=FALSE}
nsamples <- length(unique(seurat_obj$sampleid))
spatial_dimplot_width <- 6*round(sqrt(nsamples))
spatial_dimplot_height <- 6*ceiling(nsamples/round(sqrt(nsamples)))
```

```{r spatial_dimplot, fig.width=spatial_dimplot_width, fig.height=spatial_dimplot_height}
p <- SpatialDimPlot(seurat_obj, label = TRUE, label.size = 3, pt.size.factor = 2, ncol = round(sqrt(nsamples)))
ggsave('./spatial_dimplot.png',p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,dpi=300) 
ggsave('./spatial_dimplot.pdf',p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,useDingbats=FALSE)
p
```

```{r, echo=FALSE}
if(opt$cellcycle_correction_flag =="1") cellcycle_correction_flag <- TRUE else {
  cellcycle_correction_flag <- FALSE
  dimplot_width_condition <- default_width
  dimplot_height_condition <- default_height
  }
```

## Cell cycle phase

```{r dim_reduc_phase, fig.width=dimplot_width_condition, fig.height=dimplot_height_condition,eval=cellcycle_correction_flag, echo=cellcycle_correction_flag}
p <- CellDimPlot(seurat_obj,group_by = "Phase", reduction = opt$vismethod, theme = "theme_blank", highlight =TRUE, label = TRUE, label_insitu = TRUE, split_by = "condition", ncol = round(sqrt(nconditions)))
ggsave(paste0("./",opt$vismethod,'_cellcycle.png'),p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,dpi=300) 
ggsave(paste0("./",opt$vismethod,'_cellcycle.pdf'),p,width=dimplot_width_condition,height=dimplot_height_condition,limitsize = FALSE,useDingbats=FALSE)
p
```

```{r, echo=FALSE}
if(ncol(sampleinfo) >3){
  for(i in 4:ncol(sampleinfo)){
    p <- CellDimPlot(seurat_obj,group_by = colnames(sampleinfo)[i], reduction = opt$vismethod, theme = "theme_blank", highlight =TRUE, label = TRUE, label_insitu = TRUE)
    ggsave(paste0("./",opt$vismethod,'_',colnames(sampleinfo)[i],'.png'),p,width=6,height=5,dpi = 300)
    ggsave(paste0("./",opt$vismethod,'_',colnames(sampleinfo)[i],'.pdf'),p,width=6,height=5,useDingbats=F)
  }
}
```

# {-}

# Feature genes {.tabset}

## Feature gene expression

```{r, echo=FALSE}
if(file.exists(as.character(opt$feature_list))) genelist_flag <- TRUE else {
  genelist_flag <- FALSE
  featureplot_feature_width <- feature_gene_violin_width <- default_width
  featureplot_feature_height <- feature_gene_violin_height <- default_height
  }
```

```{r featureplot_feature, eval=genelist_flag, echo=genelist_flag, fig.show='hide'}
genes <- read.delim(opt$feature_list, header=F,stringsAsFactors=F)
genes <- genes$V1
if(nconditions ==1){
  p <- FeaturePlot(seurat_obj,features=genes,cols=c('grey','red'),combine = F,reduction = opt$vismethod, keep.scale = "feature", label = TRUE, pt.size = 2)
  m <- grid.arrange(grobs=p,ncol=round(sqrt(length(p))),nrow=ceiling(length(p)/round(sqrt(length(p)))))
  featureplot_feature_width <- 6*round(sqrt(length(p)))
  featureplot_feature_height <- 6*ceiling(length(p)/round(sqrt(length(p))))
} else {
  m <- FeaturePlot(seurat_obj,features=genes,cols=c('grey','red'),combine = TRUE,reduction = opt$vismethod, keep.scale = "feature", split.by = "condition", label = TRUE, pt.size = 2) & theme(legend.position = "right")
  featureplot_feature_width <- 6*nconditions
  featureplot_feature_height <- 6*ceiling(length(p)/nconditions)
}
ggsave('./feature_gene.png',m,width=featureplot_feature_width,height=featureplot_feature_height,limitsize=F,dpi=300)
ggsave('./feature_gene.pdf',m,width=featureplot_feature_width,height=featureplot_feature_height,useDingbats=F)
```

```{r featureplot_feature1, fig.width=featureplot_feature_width, fig.height=featureplot_feature_height, echo=FALSE, eval=genelist_flag}
grid.draw(m)
```

## Feature gene expression violin plot

```{r feature_gene_violin, eval=genelist_flag, echo=genelist_flag, fig.show='hide'}
p <- VlnPlot(seurat_obj,features=genes,group.by='seurat_clusters',split.by='condition',pt.size=0,combine=F)
m <- grid.arrange(grobs=p,ncol=round(sqrt(length(p))),nrow=ceiling(length(p)/round(sqrt(length(p)))))
ggsave('./feature_gene_violin.png',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,dpi=300)  
ggsave('./feature_gene_violin.pdf',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,useDingbats=F)
feature_gene_violin_width <- 12*round(sqrt(length(p)))
feature_gene_violin_height <- 6*ceiling(length(p)/round(sqrt(length(p))))
```

```{r feature_gene_violin1, fig.width=feature_gene_violin_width, fig.height=feature_gene_violin_height,echo=FALSE, eval=genelist_flag}
grid.draw(m)
```

## Feature gene expression spatial plot

```{r}
feature_gene_spatial_width <- 6*nsamples
feature_gene_spatial_height <- 6*length(genes)
```

```{r feature_gene_spatial, eval=genelist_flag, echo=genelist_flag,fig.width=feature_gene_spatial_width, fig.height=feature_gene_spatial_height}
p <- SpatialFeaturePlot(seurat_obj, features = genes, pt.size.factor = 2, keep.scale = "feature")
ggsave('./feature_gene_spatial.png',p,width=feature_gene_spatial_width,height=feature_gene_spatial_height,limitsize = F,dpi=300)  
ggsave('./feature_gene_spatial.pdf',p,width=feature_gene_spatial_width,height=feature_gene_spatial_height,limitsize = F,useDingbats=F)
p
```

# {-}

# Cluster specific genes {.tabset}

## Number of marker genes

```{r, echo=FALSE}
if(opt$marker_gene_filtered != "NA") marker_flag <- TRUE else {
  marker_flag <- FALSE
  featureplot_width <- vlnplot_width <- default_width
  featureplot_height <- vlnplot_height <- default_height
  }
```

```{r num_marker_genes_barplot, eval=marker_flag, echo=marker_flag}
marker_gene_file <- file.path(opt$marker_gene_filtered)
selectedclusters <- getSheetNames(marker_gene_file)
selectedclusters <- selectedclusters[-length(selectedclusters)]
selectedclusters <- gsub("marker_gene_","",selectedclusters)
cluster_list_filter <- lapply(1:length(selectedclusters), function(i) read.xlsx(marker_gene_file, sheet = i, rowNames = FALSE, colNames = TRUE))
df <- data.frame(cluster=factor(selectedclusters, levels = selectedclusters),
                 num=sapply(cluster_list_filter,function(i) nrow(i)))
df$num[is.na(df$num)] <- 0
gp <- ggplot(data=df) + geom_bar(aes(x=cluster,y=num,fill=cluster),position ='identity',stat='identity') +
  theme_classic() + coord_flip() + guides(fill = FALSE) +
  #scale_x_continuous(breaks = selectedclusters,labels = selectedclusters)+
  scale_y_continuous(breaks=labeling::extended(range(df$num)[1],range(df$num)[2],4),labels = abs(labeling::extended(range(df$num)[1],range(df$num)[2],4))) +
  labs(x='Cluster Number',y='Number of DEGs')
ggsave('./marker_gene_barplot.png',gp,width=6,heigh=6,dpi=300)
ggsave('./marker_gene_barplot.pdf',gp,width=6,heigh=6,useDingbats = F)
gp
```

## Top cluster marker gene feature plot

```{r marker_gene_featureplot, eval=marker_flag, echo=marker_flag, fig.show='hide'}
#genelist_top1 <- sapply(1:length(selectedclusters), function(i) {cluster_list_filter[[i]]$gene_id[1]})
genelist_top1_gs <- sapply(1:length(selectedclusters), function(i) {cluster_list_filter[[i]]$gene_name[1]})
labels <- paste0(selectedclusters,':',genelist_top1_gs)
labels <- labels[!is.na(genelist_top1_gs)]
genelist_top1_gs <- genelist_top1_gs[!is.na(genelist_top1_gs)]
#genelist_top1 <- genelist_top1[!is.na(genelist_top1)]
p <- lapply(1:length(genelist_top1_gs), function(i) FeaturePlot(seurat_obj, features = genelist_top1_gs[i],cols=c('grey','red'),reduction = opt$vismethod, label = TRUE,pt.size = 2) + labs(title = labels[i]))
m <- grid.arrange(grobs=p,ncol=round(sqrt(length(p))),nrow=ceiling(length(p)/round(sqrt(length(p)))))
ggsave('./marker_gene.png',m,width=12*round(sqrt(length(p))),height=12*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,dpi=300)
ggsave('./marker_gene.pdf',m,width=12*round(sqrt(length(p))),height=12*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,useDingbats=FALSE)
marker_gene_featureplot_width <- 12*round(sqrt(length(p)))
marker_gene_featureplot_height <- 12*ceiling(length(p)/round(sqrt(length(p))))
```

```{r marker_gene_featureplot1, fig.width=marker_gene_featureplot_width, fig.height=marker_gene_featureplot_height, eval=marker_flag, echo=FALSE}
grid.draw(m)
```

## Top cluster marker gene dotplot

```{r marker_gene_dotplot, eval=marker_flag, echo=marker_flag}
seurat_obj <- ScaleData(seurat_obj, features = rownames(seurat_obj), verbose = FALSE)

rows_data <- data.frame(genes = genelist_top1_gs,
                        clusters = gsub("cluster|:.*$","",labels))
p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", cell_type = "dot", plot_type = "heatmap", name = "Expression Level", dot_size = function(x) sum(x > 0) / length(x), dot_size_name = "Percent Expressed", add_bg = FALSE, rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, cluster_columns = FALSE, row_name_annotation = TRUE, rows_split_palette = "Set2", columns_split_palette = "Set2", columns_split_by = "seurat_clusters", layer = "scale.data")
ggsave('./marker_gene_dotplot.png',p,width=12,height=12, dpi = 300)
ggsave('./marker_gene_dotplot.pdf',p,width=12,height=12, useDingbats=FALSE)
p
```

## Top cluster marker gene violin plot

```{r marker_gene_vlnplot, eval=marker_flag, echo=marker_flag,fig.show='hide'}
p <- lapply(1:length(genelist_top1_gs), function(i) VlnPlot(seurat_obj,features=genelist_top1_gs[i],group.by='seurat_clusters',pt.size=0) + labs(title = labels[i]) + theme(legend.position = "none"))
m <- grid.arrange(grobs=p,ncol=round(sqrt(length(p))),nrow=ceiling(length(p)/round(sqrt(length(p)))))
ggsave('./marker_gene_violin.png',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,dpi=300)
ggsave('./marker_gene_violin.pdf',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,useDingbats=FALSE)
vlnplot_width <- 12*round(sqrt(length(p)))
vlnplot_height <- 6*ceiling(length(p)/round(sqrt(length(p))))
```

```{r marker_gene_vlnplot1, fig.width=vlnplot_width, fig.height=vlnplot_height, eval=marker_flag, echo=FALSE}
grid.draw(m)
```

## Top cluster marker gene spatial plot

```{r}
marker_gene_spatial_width <- 6*nsamples
marker_gene_spatial_height <- 6*length(genelist_top1_gs)
```

```{r marker_gene_spatial, eval=marker_flag, echo=marker_flag,fig.width=marker_gene_spatial_width, fig.height=marker_gene_spatial_height}
p <- SpatialFeaturePlot(seurat_obj, features = genelist_top1_gs, pt.size.factor = 2, keep.scale = "feature")
ggsave('./marker_gene_spatial.png',p,width=marker_gene_spatial_width,height=marker_gene_spatial_height,limitsize = F,dpi=300)  
ggsave('./marker_gene_spatial.pdf',p,width=marker_gene_spatial_width,height=marker_gene_spatial_height,limitsize = F,useDingbats=F)
p
```

## Top cluster marker gene heatmap

```{r marker_gene_heatmap,eval=marker_flag, echo=marker_flag}
p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", plot_type = "heatmap", name = "Expression Level", rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, row_name_annotation = TRUE, cluster_columns = FALSE, rows_split_palette = "Set2", columns_split_palette = "Set2", layer = "scale.data", columns_split_by = "seurat_clusters")

ggsave('./marker_gene_heatmap.png',p,width=8,height=12, dpi = 300)
ggsave('./marker_gene_heatmap.pdf',p,width=8,height=12,useDingbats=FALSE)
p
```

```{r figures_top5, echo=FALSE, eval=marker_flag, fig.show='hide'}
#genelist_top5 <- lapply(1:length(selectedclusters), function(i) {cluster_list_filter[[i]]$gene_id[0:min(5,nrow(cluster_list_filter[[i]]))]})
genelist_top5_gs <- lapply(1:length(selectedclusters), function(i) {
  if(nrow(cluster_list_filter[[i]]) == 0) NA else cluster_list_filter[[i]]$gene_name[0:min(5,nrow(cluster_list_filter[[i]]))]
  })
labels_top5 <- lapply(1:length(selectedclusters),function(i) paste0( selectedclusters[i],':',genelist_top5_gs[[i]]))
labels_top5 <- labels_top5[sapply(genelist_top5_gs,function(i) !any(is.na(i)))]
genelist_top5_gs <- genelist_top5_gs[sapply(genelist_top5_gs,function(i) !any(is.na(i)))]

rows_data <- data.frame(genes = unlist(genelist_top5_gs),
                        clusters = gsub("cluster|:.*$","",unlist(labels_top5)))

p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", plot_type = "heatmap", name = "Expression Level", rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, row_name_annotation = TRUE, cluster_columns = FALSE, rows_split_palette = "Set2", columns_split_palette = "Set2", layer = "scale.data", columns_split_by = "seurat_clusters")

ggsave('./marker_gene_heatmap_top5.png',p,width=8,height=12, dpi = 300)
ggsave('./marker_gene_heatmap_top5.pdf',p,width=8,height=12,useDingbats=FALSE)

p <- lapply(1:length(genelist_top5_gs),function(i) {
  p1 <- FeaturePlot(seurat_obj, features = genelist_top5_gs[[i]],combine = F,reduction = opt$vismethod, label = TRUE,pt.size = 2)
  lapply(1:length(p1),function(j) p1[[j]]+labs(title = labels_top5[[i]][j]))})
p <- unlist(p,recursive = F)
m <- matrix(NA,nrow = length(selectedclusters),ncol = 5)
m[1,1:length(genelist_top5_gs[[1]])] <- 1:length(genelist_top5_gs[[1]])
for(i in 2:length(genelist_top5_gs)){
  start <- max(m[i-1,],na.rm = T)
  m[i,1:length(genelist_top5_gs[[i]])] <- (start+1):(start+length(genelist_top5_gs[[i]]))
}
mp <- grid.arrange(grobs=p,layout_matrix=m)
ggsave('./marker_gene_top5.png',mp,width=12*5,height =12*length(genelist_top5_gs),limitsize = F,dpi=72)
ggsave('./marker_gene_top5.pdf',mp,width=12*5,height =12*length(genelist_top5_gs),limitsize = F,useDingbats = FALSE)

p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", cell_type = "dot", plot_type = "heatmap", name = "Expression Level", dot_size = function(x) sum(x > 0) / length(x), dot_size_name = "Percent Expressed", add_bg = FALSE, rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, cluster_columns = FALSE, row_name_annotation = TRUE, rows_split_palette = "Set2", columns_split_palette = "Set2", columns_split_by = "seurat_clusters", layer = "scale.data")
ggsave('./marker_gene_dotplot_top5.png',p,width=12*5,height=12,limitsize = F,dpi=72)
ggsave('./marker_gene_dotplot_top5.pdf',p,width=12*5,height=12,limitsize = F,useDingbats = FALSE)

## violin plot
p <- lapply(1:length(genelist_top5_gs),function(i){
  p1 <- VlnPlot(seurat_obj, features=genelist_top5_gs[[i]],group.by='seurat_clusters',pt.size=0,combine=F)
  lapply(1:length(p1),function(j) p1[[j]]+labs(title = labels_top5[[i]][j]) + theme(legend.position = "none"))
})
p <- unlist(p,recursive = F)
m <- matrix(NA,nrow = length(selectedclusters),ncol = 5)
m[1,1:length(genelist_top5_gs[[1]])] <- 1:length(genelist_top5_gs[[1]])
for(i in 2:length(genelist_top5_gs)){
  start <- max(m[i-1,],na.rm = T)
  m[i,1:length(genelist_top5_gs[[i]])] <- (start+1):(start+length(genelist_top5_gs[[i]]))
}
mp <- grid.arrange(grobs=p,layout_matrix=m)
ggsave('./marker_gene_violin_top5.png',mp,width=12*5,height =6*length(genelist_top5_gs),limitsize = F,dpi=72)
ggsave('./marker_gene_violin_top5.pdf',mp,width=12*5,height =6*length(genelist_top5_gs),limitsize = F,useDingbats=FALSE)
```

# {-}

# Differentially expressed genes between conditions (case vs control) {.tabset}

## Number of differential genes

```{r, echo=FALSE}
if(opt$diff_gene_filtered != "NA") {
  diff_flag = TRUE
  #opt$diff_gene_filtered <- "./diff_gene_table_filter.xlsx"
  } else {
  diff_flag = FALSE
  featureplot_width <- vlnplot_width <- default_width
  featureplot_height <- vlnplot_height <- default_height
  }
```


```{r num_diff_genes_barplot, eval=diff_flag, echo=diff_flag}
marker_gene_file <- opt$diff_gene_filtered
selectedclusters <- getSheetNames(marker_gene_file)
selectedclusters <- selectedclusters[-length(selectedclusters)]
selectedclusters <- gsub("diff_gene_","",selectedclusters)
cluster_list_filter <- lapply(1:length(selectedclusters), function(i){
  tmp <- read.xlsx(marker_gene_file, sheet = i, rowNames = FALSE, colNames = TRUE)
  idx_fc <- grep("log2FC$", colnames(tmp))
  tmp[order(abs(tmp[,idx_fc]),decreasing = TRUE),]
})
df <- data.frame(cluster=factor(rep(selectedclusters,times=2), levels = selectedclusters),
                 condition=factor(rep(c('up','down'),each=(length(cluster_list_filter))),levels = c('up','down')),
                 num=c(sapply(cluster_list_filter,function(i) sum(i[["avg_log2FC"]]>0)), -1*sapply(cluster_list_filter,function(i) sum(i[["avg_log2FC"]]<0))))
df$num[is.na(df$num)] <- 0
gp <- ggplot(data=df) + geom_bar(aes(x=cluster,y=num,fill=condition),position ='identity',stat='identity') +
  theme_classic() + coord_flip() +
  #scale_x_continuous(breaks = selectedclusters,labels = selectedclusters)+
  scale_y_continuous(breaks=labeling::extended(range(df$num)[1],range(df$num)[2],4),labels = abs(labeling::extended(range(df$num)[1],range(df$num)[2],4))) +
  labs(x='Cluster Number',y='Number of DEGs')
ggsave('./diff_gene_barplot.png',gp,width=6,heigh=6,dpi=300)
ggsave('./diff_gene_barplot.pdf',gp,width=6,heigh=6,useDingbats=FALSE)
gp
```

## Top differential gene feature plot

```{r diff_gene_featureplot,eval=diff_flag, echo=diff_flag,fig.show='hide'}
#genelist_top1 <- sapply(1:length(selectedclusters), function(i) {cluster_list_filter[[i]]$gene_id[1]})
genelist_top1_gs <- sapply(1:length(selectedclusters), function(i) {cluster_list_filter[[i]]$gene_name[1]})
labels <- paste0(selectedclusters,':',genelist_top1_gs)
labels <- labels[!is.na(genelist_top1_gs)]
genelist_top1_gs <- genelist_top1_gs[!is.na(genelist_top1_gs)]

p <- lapply(1:length(genelist_top1_gs), function(i) FeaturePlot(seurat_obj,features=genelist_top1_gs[i],cols=c('grey','red'),combine = TRUE,reduction = opt$vismethod, keep.scale = "feature", split.by = "condition", label = TRUE,pt.size = 2) + theme(legend.position = "right") + scale_y_continuous(sec.axis = dup_axis(name = labels[i])))

#m <- grid.arrange(grobs=p,ncol=1,nrow=length(genelist_top1_gs))
m <- wrap_plots(p, ncol = 1)
featureplot_width <- 6*nconditions
featureplot_height <- 6*length(genelist_top1_gs)

ggsave('./diff_gene.png',m,width=featureplot_width,height=featureplot_height,limitsize=F,dpi=300)
ggsave('./diff_gene.pdf',m,width=featureplot_width,height=featureplot_height,useDingbats=F,limitsize=F)
```

```{r diff_gene_featureplot1, fig.width=featureplot_width, fig.height=featureplot_height,eval=diff_flag, echo=FALSE}
grid.draw(m)
```

## Top differential gene dotplot

```{r diff_gene_dotplot, eval=diff_flag, echo=marker_flag}
rows_data <- data.frame(genes = genelist_top1_gs,
                        clusters = gsub("cluster|:.*$","",labels))
p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", cell_type = "dot", plot_type = "heatmap", name = "Expression Level", dot_size = function(x) sum(x > 0) / length(x), dot_size_name = "Percent Expressed", add_bg = FALSE, rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, cluster_columns = FALSE, row_name_annotation = TRUE, rows_split_palette = "Set2", columns_split_palette = "Set2", columns_split_by = "seurat_clusters",layer = "scale.data")
ggsave('./diff_gene_dotplot.png',p,width=12,height=12, dpi = 300)
ggsave('./diff_gene_dotplot.pdf',p,width=12,height=12, useDingbats=FALSE)
p
```

## Top differential gene violin plot

```{r diff_gene_vlnplot,eval=diff_flag, echo=diff_flag,fig.show='hide'}
p <- lapply(1:length(genelist_top1_gs), function(i) VlnPlot(seurat_obj,features=genelist_top1_gs[i],group.by='seurat_clusters',pt.size=0, split.by = "condition") + labs(title = labels[i]) + theme(legend.position = "right"))
m <- grid.arrange(grobs=p,ncol=round(sqrt(length(p))),nrow=ceiling(length(p)/round(sqrt(length(p)))))
ggsave('./diff_gene_violin.png',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,dpi=300)
ggsave('./diff_gene_violin.pdf',m,width=12*round(sqrt(length(p))),height=6*ceiling(length(p)/round(sqrt(length(p)))),limitsize = F,useDingbats=FALSE)
vlnplot_width <- 12*round(sqrt(length(p)))
vlnplot_height <- 6*ceiling(length(p)/round(sqrt(length(p))))
```

```{r diff_gene_vlnplot1, fig.width=vlnplot_width, fig.height=vlnplot_height,eval=diff_flag, echo=FALSE}
grid.draw(m)
```

## Top differential gene spatial plot

```{r}
diff_gene_spatial_width <- 6*nsamples
diff_gene_spatial_height <- 6*length(genelist_top1_gs)
```

```{r diff_gene_spatial, eval=diff_flag, echo=diff_flag,fig.width=diff_gene_spatial_width, fig.height=diff_gene_spatial_height}
p <- lapply(1:length(genelist_top1_gs), function(i) SpatialFeaturePlot(seurat_obj, features = genelist_top1_gs[i], pt.size.factor = 2, keep.scale = "feature") & labs(fill = labels[i]))

m <- wrap_plots(p, ncol = 1)
ggsave('./diff_gene_spatial.png',m,width=diff_gene_spatial_width,height=diff_gene_spatial_height,limitsize = F,dpi=300)  
ggsave('./diff_gene_spatial.pdf',m,width=diff_gene_spatial_width,height=diff_gene_spatial_height,limitsize = F,useDingbats=F)
m
```

## Top differential gene heatmap

```{r diff_gene_heatmap,eval=diff_flag, echo=diff_flag}
p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", plot_type = "heatmap", name = "Expression Level", rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, row_name_annotation = TRUE, cluster_columns = FALSE, rows_split_palette = "Set2", columns_split_palette = "Set2", layer = "scale.data", columns_split_by = "seurat_clusters")

ggsave('./diff_gene_heatmap.png',p,width=8,height=12, dpi = 300)
ggsave('./diff_gene_heatmap.pdf',p,width=8,height=12,useDingbats=FALSE)
p
```

```{r diff_figures_top5, echo=FALSE,eval=diff_flag, fig.show='hide'}
genelist_top5_gs <- lapply(1:length(selectedclusters), function(i) {
  if(nrow(cluster_list_filter[[i]]) == 0) NA else cluster_list_filter[[i]]$gene_name[0:min(5,nrow(cluster_list_filter[[i]]))]
  })
labels_top5 <- lapply(1:length(selectedclusters),function(i) paste0( selectedclusters[i],':',genelist_top5_gs[[i]]))
labels_top5 <- labels_top5[sapply(genelist_top5_gs,function(i) !any(is.na(i)))]
genelist_top5_gs <- genelist_top5_gs[sapply(genelist_top5_gs,function(i) !any(is.na(i)))]

rows_data <- data.frame(genes = unlist(genelist_top5_gs),
                        clusters = gsub("cluster|:.*$","",unlist(labels_top5)))

p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", plot_type = "heatmap", name = "Expression Level", rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, row_name_annotation = TRUE, cluster_columns = FALSE, rows_split_palette = "Set2", columns_split_palette = "Set2", layer = "scale.data", columns_split_by = "seurat_clusters")

ggsave('./diff_gene_heatmap_top5.png',p,width=8,height=12,dpi = 300)
ggsave('./diff_gene_heatmap_top5.pdf',p,width=8,height=12,useDingbats=FALSE)

p <- FeatureStatPlot(seurat_obj, features = rows_data$genes, ident = "condition", cell_type = "dot", plot_type = "heatmap", name = "Expression Level", dot_size = function(x) sum(x > 0) / length(x), dot_size_name = "Percent Expressed", add_bg = FALSE, rows_data = rows_data, show_row_names = TRUE, rows_split_by = "clusters", cluster_rows = FALSE, cluster_columns = FALSE, row_name_annotation = TRUE, rows_split_palette = "Set2", columns_split_palette = "Set2", columns_split_by = "seurat_clusters", layer = "scale.data")

ggsave('./diff_gene_dotplot_top5.png',p,width=12*5,height=12,limitsize = F,dpi=72)
ggsave('./diff_gene_dotplot_top5.pdf',p,width=12*5,height=12,limitsize = F,useDingbats = FALSE)

## violin plot
p <- lapply(1:length(genelist_top5_gs),function(i){
  p1 <- VlnPlot(seurat_obj, features=genelist_top5_gs[[i]],group.by='seurat_clusters',pt.size=0,combine=F,split.by = "condition")
  lapply(1:length(p1),function(j) p1[[j]]+labs(title = labels_top5[[i]][j]) + theme(legend.position = "none"))
})
p <- unlist(p,recursive = F)
m <- matrix(NA,nrow = length(selectedclusters),ncol = 5)
m[1,1:length(genelist_top5_gs[[1]])] <- 1:length(genelist_top5_gs[[1]])
for(i in 2:length(genelist_top5_gs)){
  start <- max(m[i-1,],na.rm = T)
  m[i,1:length(genelist_top5_gs[[i]])] <- (start+1):(start+length(genelist_top5_gs[[i]]))
}
mp <- grid.arrange(grobs=p,layout_matrix=m)
ggsave('./diff_gene_violin_top5.png',mp,width=12*5,height =6*length(genelist_top5_gs),limitsize = F,dpi=72)
ggsave('./diff_gene_violin_top5.pdf',mp,width=12*5,height =6*length(genelist_top5_gs),limitsize = F,useDingbats=FALSE)
```

# {-}

# Access to results files

Detailed QC results: [QC](../qc/)

Cell number by cluster summary: [tables](./tables/)

Top 5 marker genes for each cluster, if applicable:

* [feature plot](./figures/marker_gene_top5.png)
* [dot plot](./figures/marker_gene_dotplot_top5.png)
* [violin plot](./figures/marker_gene_violin_top5.png)
* [heatmap](./figures/marker_gene_heatmap_top5.png)

For marker gene list table: [marker_genes](./gene_table/marker_gene_table.xlsx)

For filtered marker gene list table based on supplied cutoffs: [marker_genes_filter](./gene_table/marker_gene_table_filter.xlsx)

Top 5 DEGs for each cluster (case vs control), if applicable:

* [feature plot](./figures/diff_gene_top5.png)
* [dot plot](./figures/diff_gene_dotplot_top5.png)
* [violin plot](./figures_diff_gene_violin_top5.png)
* [heatmap](./figures/diff_gene_heatmap_top5.png)

For DEGs within each cluster (case vs control): [diff_genes](./gene_table/diff_gene_table.xlsx)

For filtered DEGs within each cluster (case vs control): [diff_genes_filter.xlsx](./gene_table/diff_table_filter.xlsx)

seurat object: [.rds file](./)

on disk matrix (if applicable): [on_disk_mat](./on_disk_mat)