---
title: "QC report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
```

```{r load_library}
### load libraries
### matrixStats v1.2 and above seems to cause issues for glmGamPoi
renv::load(Sys.getenv("PROJECT_DIR"))
library(matrixStats)
library(SoupX)
library(scDblFinder)
library(ggplot2)
library(Seurat)
library(patchwork)
options(stringsAsFactors = F, future.globals.maxSize=50*1024^3)
opt <- params$opt
```

```{r perform_qc, echo=FALSE}
source(file.path(opt$workflowpath, "scripts/helper_functions.r"))
seurat_obj <- qcsample_scRNAseq(
    sampleid = opt$sampleid,
    condition = opt$condition,
    secondary_output = opt$secondary_output,
    stage = "QC",
    workflowpath = opt$workflowpath,
    ambient_RNA_removal_flag = opt$ambient_RNA_removal_flag,
    doublet_removal_flag = opt$doublet_removal_flag,
    adaptive_cutoff_flag = opt$adaptive_cutoff_flag,
    nCount_cutoff = opt$nCount_cutoff,
    nFeature_cutoff = opt$nFeature_cutoff,
    mt_cutoff = opt$mt_cutoff,
    hb_cutoff = opt$hb_cutoff,
    nCell_cutoff = opt$nCell_cutoff
)
```

# QC Steps

1. Ambient RNA correction using [SoupX](https://github.com/constantAmateur/SoupX) (if enabled)
2. Doublet identification using [scDblFinder](https://github.com/plger/scDblFinder)
3. Identification of cells with low number of total UMI counts (`r opt$nCount_cutoff`)
4. Identification of cells with low number of detectable features/genes (`r opt$nFeature_cutoff`)
5. Identification of cells with high percentage of mitochondrial concentration (`r opt$mt_cutoff`), indicating low quality/apoptotic cells
6. Identification of cells with high percentage of hemoglobin concentration (`r opt$hb_cutoff`), indicating red blood cell contamination
7. Identification of genes with low number of expressing cells (`r opt$nCell_cutoff`), indicating lowly expressed genes

Adaptive cutoff will be used for step 3-7 if enabled. Adaptive cutoff is more stringent, and sample-specific depending on distribution of metrics.

# QC threshold

```{r}
opt_postQC <- read.delim(paste0(opt$sampleid,"_opt_postQC.txt"), header = TRUE)
qc_table <- data.frame(metric=c("ambient_RNA_removal_flag", "doublet_removal_flag", "nCount_cutoff", "nFeature_cutoff","mt_cutoff", "hb_cutoff", "nCell_cutoff","adaptive_cutoff_flag"),
                       description=c("0 or 1 to indicate whether to perform ambient RNA removal/correction",
                                     "0 or 1 to indicate whether to perfrom doublet removal",
                                     "Cutoff for total number of UMI counts",
                                     "Cutoff for total number of detectable genes/features",
                                     "Cutoff for percentage of mitochondria concentration",
                                     "Cutoff for percentage of hemoglobin concentration",
                                     "Cutoff for number of cells with expression for feature/gene",
                                     "0 or 1 to indicate whether to apply adaptive cutoff idenfication based on IQR"),
                       cutoff_suppied=c(opt$ambient_RNA_removal_flag, opt$doublet_removal_flag, opt$nCount_cutoff, opt$nFeature_cutoff, opt$mt_cutoff, opt$hb_cutoff, opt$nCell_cutoff, opt$adaptive_cutoff_flag),
                       cutoff_used=c(opt_postQC$ambient_RNA_removal_flag, opt_postQC$doublet_removal_flag, opt_postQC$nCount_cutoff, opt_postQC$nFeature_cutoff, opt_postQC$mt_cutoff, opt_postQC$hb_cutoff, opt_postQC$nCell_cutoff, opt_postQC$adaptive_cutoff_flag))
knitr::kable(qc_table, caption = "A list of QC metrics used")
```

# Distribution of metrics {.tabset}

## Violin plot

```{r, fig.width=12, fig.height=5}
cell_metadata <- read.delim(paste0(opt$sampleid,"_qc_metrics_cells.txt"), header = TRUE)
if(length(unique(cell_metadata$scDblFinder.class)) > 1) cell_metadata$scDblFinder.class <- factor(cell_metadata$scDblFinder.class, levels = c("singlet", "doublet"))
gps <- lapply(c("nCount_RNA", "nFeature_RNA","percent.mt", "percent.hb"), function(i) {
  ggplot() + geom_violin(aes(x = opt$sampleid, y = cell_metadata[[i]]), fill = "#F8766D") + geom_jitter(aes(x = opt$sampleid, y = cell_metadata[[i]]), size=1, height = 0) + labs(x = "", y = "", title = i) + theme_classic() + theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12), axis.text.x = element_text(size = 10)) 
})
p <- wrap_plots(gps, ncol = 4)
ggsave('qc_violinplot.png',p,width=12,height=5)
p
```

## Scatter plot

```{r, fig.width=12, fig.height=12}
plot1 <- ggplot(cell_metadata, aes(x = nCount_RNA, y = percent.mt)) + geom_point(aes(shape = scDblFinder.class)) + theme_classic() + scale_shape_manual(values = c(1, 4)) + facet_wrap(~scDblFinder.class, ncol = 2)
  plot2 <- ggplot() + geom_point(aes(x = nCount_RNA, y = nFeature_RNA,shape = scDblFinder.class), data = cell_metadata) + theme_classic() + scale_shape_manual(values = c(1, 4)) + facet_wrap(~scDblFinder.class, ncol = 2)
p <- plot1 + plot2 + plot_layout(ncol=1)
ggsave("qc_scatterplot.png",p,width=12,height=12)
p
```

# {-}

# Summary

```{r}
qc_summary <- read.delim(paste0(opt$sampleid,'_qc_metrics_summary.txt'),header = TRUE)
```

The total number of cells before QC is `r nrow(cell_metadata)`. Ambient RNA rate is estimated to be `r round(unique(cell_metadata$soupX.rho)*100,2)`%. `r sum(cell_metadata$scDblFinder.class == "doublet")` # of cells were predicted to be doublets (`r round(sum(cell_metadata$scDblFinder.class == "doublet")/nrow(cell_metadata)*100,2)`% doublet rate). With defined cutoffs, `r qc_summary["hb_cutoff"]` # of cells pass all cutoffs, and `r qc_summary["total_genes"]` # of genes pass the cutoff. 

```{r, fig.width=12, fig.height=8}
qc_summary1 <- data.frame(metric = factor(names(qc_summary[-c(1,ncol(qc_summary))]), levels = names(qc_summary)[(length(qc_summary)-1):2]),
                            value = as.numeric(qc_summary[-c(1,ncol(qc_summary))]))
gp <- ggplot(aes(x=metric, y = value), data = qc_summary1) + geom_col(aes(fill=metric)) + theme_classic() + geom_text(aes(label = value)) + coord_flip() + labs(x = "", y = "# of cells passing QC") + guides(fill="none") + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12, face = "bold"))
ggsave('qc_summary.png',gp,width=12,height=8)
gp
```