---
title: "QC report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r load_library}
### load libraries
## use v1.1 for now to avoid error from sctransform
renv::load(Sys.getenv("PROJECT_DIR"))
library(matrixStats)
library(ggplot2)
library(Seurat)
library(patchwork)
options(stringsAsFactors = F, future.globals.maxSize=50*1024^3)
opt <- params$opt
```

```{r perform_qc, echo=FALSE}
source(file.path(opt$workflowpath, "scripts/helper_functions.r"))
seurat_obj <- qcsample_Visium(
    sampleid = opt$sampleid,
    condition = opt$condition,
    secondary_output = opt$secondary_output,
    stage = "QC",
    workflowpath = opt$workflowpath,
    adaptive_cutoff_flag = opt$adaptive_cutoff_flag,
    nCount_cutoff = opt$nCount_cutoff,
    nFeature_cutoff = opt$nFeature_cutoff,
    mt_cutoff = opt$mt_cutoff,
    hb_cutoff = opt$hb_cutoff,
    nCell_cutoff = opt$nCell_cutoff,
)
```

# QC Steps

1. Identification of spots with low number of total UMI counts (`r opt$nCount_cutoff`)
2. Identification of spots with low number of detectable features/genes (`r opt$nFeature_cutoff`)
3. Identification of spots with high percentage of mitochondrial concentration (`r opt$mt_cutoff`), indicating low quality/apoptotic spots
4. Identification of spots with high percentage of hemoglobin concentration (`r opt$hb_cutoff`), indicating red blood cell contamination
5. Identification of genes with low number of expressing spots (`r opt$nCell_cutoff`), indicating lowly expressed genes

Adaptive cutoff will be used if enabled. Adaptive cutoff is more stringent, and sample-specific depending on distribution of metrics.

# QC threshold

```{r}
opt_postQC <- read.delim(paste0(opt$sampleid,"_opt_postQC.txt"), header = TRUE)
qc_table <- data.frame(metric=c("nCount_cutoff", "nFeature_cutoff","mt_cutoff", "hb_cutoff", "nCell_cutoff","adaptive_cutoff_flag"),
                       description=c("Cutoff for total number of UMI counts",
                                     "Cutoff for total number of detectable genes/features",
                                     "Cutoff for percentage of mitochondria concentration",
                                     "Cutoff for percentage of hemoglobin concentration",
                                     "Cutoff for number of spots with expression for feature/gene",
                                     "0 or 1 to indicate whether to apply adaptive cutoff idenfication based on IQR"),
                cutoff_suppied=c(opt$nCount_cutoff, opt$nFeature_cutoff, opt$mt_cutoff, opt$hb_cutoff, opt$nCell_cutoff, opt$adaptive_cutoff_flag),
              cutoff_used=c(opt_postQC$nCount_cutoff, opt_postQC$nFeature_cutoff, opt_postQC$mt_cutoff, opt_postQC$hb_cutoff, opt_postQC$nCell_cutoff, opt_postQC$adaptive_cutoff_flag))
knitr::kable(qc_table, caption = "A list of QC metrics used")
```

# Distribution of metrics {.tabset}

## Violin plot

```{r, fig.width=12, fig.height=5}
cell_metadata <- read.delim(paste0(opt$sampleid,"_qc_metrics_cells.txt"), header = TRUE)
gps <- lapply(c("nCount_Spatial", "nFeature_Spatial","percent.mt", "percent.hb"), function(i) {
  ggplot() + geom_violin(aes(x = opt$sampleid, y = cell_metadata[[i]]), fill = "#F8766D") + geom_jitter(aes(x = opt$sampleid, y = cell_metadata[[i]]), size=1, height = 0) + labs(x = "", y = "", title = i) + theme_classic() + theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 12), axis.text.x = element_text(size = 10)) 
})
p <- wrap_plots(gps, ncol = 4)
ggsave('qc_violinplot.png',p,width=12,height=5)
p
```

## Scatter plot

```{r, fig.width=12, fig.height=6}
plot1 <- ggplot(seurat_obj@meta.data, aes(x = nCount_Spatial, y = percent.mt)) + geom_point() + theme_classic()
plot2 <- ggplot() + geom_point(aes(x = nCount_Spatial, y = nFeature_Spatial), data = seurat_obj@meta.data) + theme_classic()
p <- plot1 + plot2 + plot_layout(ncol=2)
ggsave("qc_scatterplot.png",p,width=12,height=6)
p
```

## Spatial plot

```{r, fig.width=18, fig.height=5}
idx <- which(seurat_obj$nCount_Spatial >= opt_postQC$nCount_cutoff & seurat_obj$nFeature_Spatial >= opt_postQC$nFeature_cutoff & seurat_obj$percent.mt <= opt_postQC$mt_cutoff & seurat_obj$percent.hb <= opt_postQC$hb_cutoff)
gene_summary <- read.delim(paste0(opt$sampleid,'_qc_metrics_genes.txt'),header = TRUE)
idx1 <- which(gene_summary$nCell > opt_postQC$nCell_cutoff)
seurat_obj_filter <- subset(seurat_obj, cells = idx, features = idx1)
plot1 <- SpatialFeaturePlot(seurat_obj, features = "nCount_Spatial",pt.size.factor = 2) + ggtitle('before filter') + theme(legend.position = "right",plot.title = element_text(hjust = 0.5))
plot2 <- SpatialFeaturePlot(seurat_obj_filter, features = "nCount_Spatial",pt.size.factor = 2) + ggtitle(paste0('after filter',' spots removed (',ncol(seurat_obj)-length(idx),')',' genes removed (',nrow(seurat_obj)-length(idx1),')')) + theme(legend.position = "right",plot.title = element_text(hjust = 0.5))
if(length(idx) < ncol(seurat_obj)) cols.highlight <- c("red", "grey") else cols.highlight <- "grey"
plot3 <- SpatialDimPlot(seurat_obj,cells.highlight = colnames(seurat_obj)[-idx],pt.size.factor = 2, cols.highlight = cols.highlight) + ggtitle('Spots removed') + theme(legend.position = "none",plot.title = element_text(hjust = 0.5))
#p <- CombinePlots(plots = list(plot1, plot2))
p <- plot1 + plot2 + plot3 + plot_layout(ncol=3)
ggsave("qc_spatialplot.png",p,width=18,height=5)
p
```

# {-}

# Summary

```{r}
qc_summary <- read.delim(paste0(opt$sampleid,'_qc_metrics_summary.txt'),header = TRUE)
```

The total number of spots before QC is `r ncol(seurat_obj)`. With defined cutoffs, `r as.numeric(qc_summary[colnames(qc_summary) == "hb_cutoff"])` # of spots pass all cutoffs, and `r format(as.numeric(qc_summary[colnames(qc_summary) == "total_genes"]), scientific = FALSE)` # of genes pass the cutoff. 

```{r, fig.width=12, fig.height=8}
qc_summary1 <- data.frame(metric = factor(names(qc_summary[-c(1,ncol(qc_summary))]), levels = names(qc_summary)[(length(qc_summary)-1):2]),
                            value = as.numeric(qc_summary[-c(1,ncol(qc_summary))]))
gp <- ggplot(aes(x=metric, y = value), data = qc_summary1) + geom_col(aes(fill=metric)) + theme_classic() + geom_text(aes(label = value)) + coord_flip() + labs(x = "", y = "# of spots passing QC") + guides(fill="none") + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12, face = "bold"))
ggsave('qc_summary.png',gp,width=12,height=8)
gp
```