---
title: "QC report summary"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: false
    lightbox: false
    gallery: false
    highlight: tango
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r load_library}
### load libraries
renv::load(Sys.getenv("PROJECT_DIR"))
library(ggplot2)
library(patchwork)
library(tidyr)
library(plotly)
library(DT)
options(stringsAsFactors = F)
```

```{r summarize_metrics}
opt <- params$opt
qc_metrics_cells_paths <- list.files("./", pattern = "*._qc_metrics_cells.txt")
qc_list <- lapply(qc_metrics_cells_paths, function(i) read.delim(i, header = TRUE))
ncells <- sapply(qc_list, function(i) nrow(i))
opt_postqc_paths <- list.files("./", pattern = "*._opt_postQC.txt")
opt_postqc_list <- lapply(opt_postqc_paths, function(i) read.delim(i, header = TRUE))
opt_postqc_list <- opt_postqc_list[match(sapply(qc_list, function(i) unique(i$sampleid)),sapply(opt_postqc_list, function(i) i$sampleid))]
sample_metrics_pct <- do.call(rbind,lapply(qc_list, function(i) {
  data.frame(sampleid = unique(i$sampleid),
             nCount_Spatial = sum(i$nCount_Spatial < opt$nCount_cutoff)/nrow(i),
             nFeature_Spatial = sum(i$nFeature_Spatial < opt$nFeature_cutoff)/nrow(i),
             percent.mt = sum(i$percent.mt > opt$mt_cutoff)/nrow(i),
             percent.hb = sum(i$percent.hb > opt$hb_cutoff)/nrow(i))
}))
sample_metrics_long <- pivot_longer(as.data.frame(sample_metrics_pct), cols = colnames(sample_metrics_pct)[-1])
sample_metrics_long$value <- sample_metrics_long$value*100
sample_metrics_long$name <- factor(sample_metrics_long$name, levels = colnames(sample_metrics_pct)[-1])

sample_metrics_pct_postQC <- do.call(rbind,lapply(1:length(qc_list), function(i) {
  data.frame(sampleid = unique(qc_list[[i]]$sampleid),
             nCount_Spatial = sum(qc_list[[i]]$nCount_Spatial < opt_postqc_list[[i]]$nCount_cutoff)/nrow(qc_list[[i]]),
             nFeature_Spatial = sum(qc_list[[i]]$nFeature_Spatial < opt_postqc_list[[i]]$nFeature_cutoff)/nrow(qc_list[[i]]),
             percent.mt = sum(qc_list[[i]]$percent.mt > opt_postqc_list[[i]]$mt_cutoff)/nrow(qc_list[[i]]),
             percent.hb = sum(qc_list[[i]]$percent.hb > opt_postqc_list[[i]]$hb_cutoff)/nrow(qc_list[[i]]))
}))
sample_metrics_long_postQC <- pivot_longer(as.data.frame(sample_metrics_pct_postQC), cols = colnames(sample_metrics_pct_postQC)[-1])
sample_metrics_long_postQC$value <- sample_metrics_long_postQC$value*100
sample_metrics_long_postQC$name <- factor(sample_metrics_long_postQC$name, levels = colnames(sample_metrics_pct_postQC))

sample_metrics_median <- do.call(rbind,lapply(qc_list, function(i) {
  apply(i[,c("nCount_Spatial", "nFeature_Spatial", "percent.mt", "percent.hb")], 2, median)
}))
sample_metrics_median <- as.data.frame(sample_metrics_median)
sample_metrics_median$sampleid <- sapply(qc_list, function(i) unique(i$sampleid))
sample_metrics_median_long <- pivot_longer(as.data.frame(sample_metrics_median), cols = colnames(sample_metrics_median)[-ncol(sample_metrics_median)])
for(i in c("nCount_Spatial", "percent.mt", "percent.hb")) sample_metrics_median[,i] <- round(sample_metrics_median[,i], digits = 2)

qc_metrics_summary_paths <- list.files("./", pattern = "*._qc_metrics_summary.txt")
qc_summary_list <- do.call(rbind, lapply(qc_metrics_summary_paths, function(i) read.delim(i, header = TRUE)))
qc_summary_list_long <- pivot_longer(qc_summary_list, cols = colnames(qc_summary_list)[-1])

median_umi_paths <- list.files("./", pattern = "*._median_umi.txt")
min_median_umi <- min(do.call(c, lapply(median_umi_paths, function(i) read.delim(i, header = FALSE)$V1)))
write.table(min_median_umi, "min_median_umi.txt",row.names = F, col.names = F, sep = "\t", quote = F)
```

# Data summary

There are a total of `r length(qc_metrics_summary_paths)` samples (total `r sum(ncells)` # of spots). The range of # of spots is `r paste0(min(ncells), '-', max(ncells))`. The range of median UMI counts is `r paste0(min(sample_metrics_median[,"nCount_Spatial"]),"-",max(sample_metrics_median[,"nCount_Spatial"]))`. The range of median detectable features is `r paste0(min(sample_metrics_median[,"nFeature_Spatial"]),"-",max(sample_metrics_median[,"nFeature_Spatial"]))`. The range of median mitochondrial (MT) concentration is `r paste0(min(sample_metrics_median[,"percent.mt"]),"-",max(sample_metrics_median[,"percent.mt"]), "%")`. The range of median hemoglobin (Hb) concentration is `r paste0(min(sample_metrics_median[,"percent.hb"]),"-",max(sample_metrics_median[,"percent.hb"]), "%")`. The range of sample-level total number of captured genes is `r paste0(min(qc_summary_list$total_genes), "-", max(qc_summary_list$total_genes))`

# Distribution of QC metrics {.tabset}

## Metric median values

```{r}
#### plotly duplicate outlier jitter points
#### https://github.com/plotly/plotly.R/issues/1114#issuecomment-1803541613
remove_boxplot_outliers <- function(fig){
  stopifnot("plotly" %in% class(fig))
  fig$x$data <- lapply(
    fig$x$data,
    \(i){
      if(i$type != "box") return(i)
      i$marker = list(opacity = 0)
      i$hoverinfo = "none"
      i
    }
  )
  fig
}
gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = sample_metrics_median_long) + geom_boxplot() + theme_classic() + labs(x = "", y = "value") + geom_point(position = position_jitter(height = 0)) + facet_wrap(~name, nrow = 2,scales="free") + theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
remove_boxplot_outliers(ggplotly(gp))
```

## Total number of genes captured per sample

```{r}
gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = qc_summary_list_long[qc_summary_list_long$name %in% c("total_genes"),]) + geom_boxplot() + theme_classic() + labs(x = "", y = "# of genes") + geom_point(position = position_jitter(height = 0)) + theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
remove_boxplot_outliers(ggplotly(gp))
```

# {-}

# Thresholds for QC metrics {.tabset}

## Distribution of thresholds

```{r}
opt_postqc_list <- do.call(rbind, opt_postqc_list)
col_sel <- c("sampleid","mt_cutoff", "hb_cutoff", "nFeature_cutoff", "nCount_cutoff", "nCell_cutoff")
opt_postqc_list_long <- pivot_longer(opt_postqc_list[,col_sel], cols = col_sel[-1])
gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = opt_postqc_list_long) + geom_boxplot() + theme_classic() + labs(x = "", y = "value") + geom_point(position = position_jitter(height = 0)) + facet_wrap(~name, nrow = 2,scales="free") + theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank())
remove_boxplot_outliers(ggplotly(gp))
```

## Pct of spots failing each QC metric with supplied cutoff

```{r}
gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = sample_metrics_long) + geom_boxplot() + theme_classic() + labs(x = "", y = "% of spots fail a certian metric") + geom_point(position = position_jitter(height = 0)) + theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
remove_boxplot_outliers(ggplotly(gp))
```

## Pct of spots failing each QC metric with adaptive cutoff

```{r}
if(opt$adaptive_cutoff == "1"){
  gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = sample_metrics_long_postQC) + geom_boxplot() + theme_classic() + labs(x = "", y = "% of spots fail a certian metric") + geom_point(position = position_jitter(height = 0)) + theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
remove_boxplot_outliers(ggplotly(gp))
} else print("adaptive cutoff is disabled")
```

# {-}

# QC summary {.tabset}

## box plot

```{r}
qc_summary_list_long <- qc_summary_list_long[qc_summary_list_long$name != "total_genes",]
qc_summary_list_long$name <- factor(qc_summary_list_long$name, levels = rev(colnames(qc_summary_list)))
gp <- ggplot(aes(x= name, y = value, color = name,label = sampleid), data = qc_summary_list_long) + geom_boxplot() + geom_point(position = position_jitter(height = 0)) + theme_classic() + coord_flip() + labs(x = "", y = "# of spots passing QC") + guides(fill="none") + theme(axis.text = element_text(size = 10), axis.title = element_text(size = 12, face = "bold"))
remove_boxplot_outliers(ggplotly(gp))
```

## table

```{r}
datatable(qc_summary_list, rownames = TRUE)
```

# {-}